\documentclass[11pt]{article}

% ---------------------------------------------------------------------------
% Formal paper: an independent transformer oracle for auditing native_decide.
% Build:  pdflatex transformer-oracle-lean-audit.tex   (twice, for references)
% Only standard packages are used so it compiles out of the box.
% ---------------------------------------------------------------------------
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[hidelinks]{hyperref}

\lstset{basicstyle=\ttfamily\small,columns=fullflexible,breaklines=true,
        keywordstyle=\color{blue!60!black},commentstyle=\color{green!40!black},
        frame=single,framesep=3pt,xleftmargin=6pt}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

\newcommand{\Word}{\mathbb{W}}
\newcommand{\Reg}{\mathsf{Reg}}
\newcommand{\Addr}{\mathsf{Addr}}
\newcommand{\Byte}{\mathsf{Byte}}
\newcommand{\Bool}{\mathbb{B}}
\newcommand{\step}{\longrightarrow}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\bp}{\{-1,+1\}}
% double square brackets without depending on stmaryrd (keeps the build package-light)
\providecommand{\llbracket}{\mathopen{[\![}}
\providecommand{\rrbracket}{\mathclose{]\!]}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\eval}{\mathcal{E}}

\title{\bfseries An Independent Transformer Oracle for Auditing\\ Lean 4's \texttt{native\_decide}}
\author{Transformer-Oracle}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Lean~4's \texttt{native\_decide} tactic discharges a decidable proposition by \emph{compiling} its
\texttt{Decidable} instance to native code, executing it, and trusting the result. This is fast, but
it enlarges the trusted computing base to include the Lean$\to$C compiler, every
\texttt{@[extern]}/\texttt{@[implemented\_by]} fast path the procedure touches, and the physical CPU.
We present Transformer-Oracle, a system that compiles ordinary programs into the weights of a transformer and
executes them there, and we use it as an \emph{independent execution oracle} to audit
\texttt{native\_decide}. We give an operational semantics for the target ISA (NISA), show how a
weight-constructed transformer realises its one-step relation in a bipolar $\bp$ tensor encoding, and
describe a freestanding mini Lean runtime that lets the emitted C run through the pipeline. We then
formalise the audit as a \emph{sound but incomplete falsifier}: a disagreement between the compiled
procedure and the independent evaluators is a proof of a fault, whereas agreement is only evidence.
As the method's flagship result we highlight a confirmed soundness bug in the Batteries library's
\texttt{Vector.scanr} fast path---an off-by-one that lets \texttt{native\_decide} prove
\texttt{False}, reported upstream as Batteries issue~\#1924~\cite{batteries1924}---and, applying the
same methodology, defects in unrelated mature libraries. We close with the threat model, including
silent hardware data corruption.
\end{abstract}

\section{Introduction}

A proof assistant is only as trustworthy as the code that decides its proof obligations. Lean~4~\cite{demoura2021}
offers two tactics for closing a \emph{decidable} goal $P$:
\begin{itemize}[nosep]
  \item \texttt{decide}, which evaluates the \texttt{Decidable P} instance \emph{inside the kernel};
        the only trusted component is the kernel's reduction machine;
  \item \texttt{native\_decide}, which \emph{compiles} the instance to C, then to native code, runs
        it, and accepts $P$ if the program returns \texttt{isTrue}.
\end{itemize}
\texttt{native\_decide} is often orders of magnitude faster, which makes it indispensable for
verification tasks that reduce to large finite computations. The cost is epistemic: correctness of
the proof now depends on the compiler, on any hand-written fast paths substituted for reference
definitions at runtime, and on the hardware. These dependencies are exactly the ones discharged by
Lean's axioms \texttt{ofReduceBool} and \texttt{trustCompiler}. If any of them fails for the specific
computation, \texttt{native\_decide} can accept a false proposition without any indication.

The contribution of this paper is a practical, \emph{independent} way to check what
\texttt{native\_decide} actually computed. Transformer-Oracle lowers C (including the C that Lean emits) to a
small register machine, NISA, and executes NISA on a transformer whose weights are \emph{constructed},
not trained, so that one forward pass performs one instruction. Constructing rather than learning
transformer weights to realise a program is the premise of \emph{compiled transformers}---RASP and its
compiler Tracr, and looped transformers used as programmable computers~\cite{weiss2021,lindner2023,giannou2023};
we take it further, compiling an entire C toolchain's output down to a transformer and using the result
as a trusted-independent executor. Because that executor shares no code with a conventional CPU --- it
computes in a bipolar $\bp$ encoding as dense tensor algebra~\cite{vaswani2017} --- running the same
decision procedure on it, and cross-checking against a separate mathematical oracle, gives a
differential test whose disagreements localise a genuine fault.

\paragraph{The governing principle.} Following Dijkstra~\cite{dijkstra1970}, testing can reveal the presence of bugs but
not their absence. Re-running a procedure and obtaining the \emph{same} answer proves nothing: the two
runs may share a defect. Obtaining a \emph{different} answer is a falsifier: it proves at least one
path is wrong (\S\ref{sec:audit}). Every design choice below serves to make disagreements meaningful
and to maximise the independence of the paths being compared.

\section{Background: \texttt{native\_decide} and its trust surface}\label{sec:background}

Let $P$ be a proposition with an instance $d : \texttt{Decidable } P$. Reducing $d$ yields either
\texttt{isTrue h} (with $h : P$) or \texttt{isFalse h}. \texttt{native\_decide} works by evaluating
the boolean \texttt{decide}~$P$ through Lean's \emph{compiler} rather than its kernel and appealing to
\[
  \texttt{ofReduceBool} : (\texttt{decide } P \leadsto_{\text{native}} \texttt{true}) \to P,
\]
where $\leadsto_{\text{native}}$ denotes evaluation of the compiled program. The trusted surface of a
single \texttt{native\_decide} call is therefore the set
\[
  \mathcal{T}(P) = \underbrace{\mathcal{C}}_{\text{Lean}\to\text{C}\to\text{native}}
                 \;\cup\; \underbrace{\mathcal{F}(d)}_{\text{fast paths reached by }d}
                 \;\cup\; \underbrace{\mathcal{H}}_{\text{CPU}} .
\]
The most commonly implicated component is $\mathcal{F}(d)$: standard-library functions carry
\texttt{@[extern]} or \texttt{@[implemented\_by]} annotations that replace the reference Lean
definition with hand-written C or a faster Lean routine during compiled execution. A discrepancy
between a fast path and its reference definition at an edge case is invisible to \texttt{decide} (which
runs the reference) but is exactly what \texttt{native\_decide} runs. Our audit targets $\mathcal{T}(P)$
as a whole by re-executing $d$ on machinery that shares none of $\mathcal{C}$, $\mathcal{F}(d)$, or
$\mathcal{H}$.

\section{The NISA abstract machine}\label{sec:nisa}

NISA is a 40-opcode, 32-bit, RISC-V-derived instruction set. We give its small-step semantics.

\begin{definition}[Words and stores]
Let $\Word = \Z/2^{32}\Z$ be the set of machine words, $\Reg = \{0,\dots,31\}$ the register indices,
$\Byte = \Z/2^{8}\Z$, and $\Addr = \{0,\dots,N-1\}$ a byte-addressed memory of size $N$. A
\emph{register file} is $R : \Reg \to \Word$ and a \emph{memory} is $M : \Addr \to \Byte$. Multi-byte
accesses are little-endian: $\mathrm{wd}_M(a) = \sum_{i=0}^{3} M(a{+}i)\,2^{8i} \in \Word$.
\end{definition}

\begin{definition}[Instructions and programs]
An instruction is a tuple $(\mathit{op}, a, b, c)$ with $\mathit{op}$ one of the 40 opcodes and
$a,b,c$ operands (register indices or immediates, per opcode). A program is a finite sequence
$P = \langle I_0, \dots, I_{n-1}\rangle$. A \emph{configuration} is $\sigma = (pc, R, M)$ with
$pc \in \N$.
\end{definition}

\begin{definition}[One-step relation]\label{def:step}
The transition $\sigma \step_P \sigma'$ is defined by cases on $I_{pc} = (\mathit{op},a,b,c)$. Writing
$R'=R[a \mapsto v]$ for the register update and $pc'=pc{+}1$ unless stated otherwise, representative
rules are:
\[
\begin{array}{llll}
\textsc{Add}   & v = R(b) + R(c) \bmod 2^{32}, & (pc,R,M)\step(pc',R',M) \\
\textsc{Mul}   & v = R(b)\cdot R(c) \bmod 2^{32}, & (pc,R,M)\step(pc',R',M) \\
\textsc{SltU}  & v = [\,R(b) < R(c)\,],          & (pc,R,M)\step(pc',R',M) \\
\textsc{Load}  & v = \mathrm{wd}_M(R(b){+}c),     & (pc,R,M)\step(pc',R',M) \\
\textsc{Store} & M' = M[\,R(b){+}c \mapsto R(a)\,], & (pc,R,M)\step(pc,R,M') \\
\textsc{Bne}   & pc' = \begin{cases} c & R(a)\neq R(b)\\ pc{+}1 & \text{else}\end{cases}, & (pc,R,M)\step(pc',R,M)\\
\textsc{Halt}  & \multicolumn{2}{l}{\text{no successor (terminal).}}
\end{array}
\]
Signed opcodes ($\textsc{Slt}, \textsc{Blt}, \textsc{Div}, \textsc{Sra}, \dots$) interpret operands via
the two's-complement bijection $\iota:\Word\to\{-2^{31},\dots,2^{31}{-}1\}$. The remaining opcodes
(bitwise, shifts, immediates, byte/half memory, high-multiply $\textsc{MulH*}$, jumps) are analogous.
\end{definition}

\begin{definition}[Whole-program evaluation]
Let $\step_P^{\,*}$ be the reflexive-transitive closure of $\step_P$. For an initial configuration
$\sigma_0=(0,R_0,M_0)$, if $\sigma_0 \step_P^{\,*} (pc,R,M)$ with $I_{pc}=\textsc{Halt}$, the machine
\emph{halts} and its result is $\sem{P}(\sigma_0) = R(10)$ (the ABI return register \texttt{a0}). If no
such terminal configuration is reachable within a cycle budget $B$, evaluation is $\bot$ (timeout).
\end{definition}

\section{The bipolar transformer executor}\label{sec:transformer}

We now realise $\step_P$ as a single transformer layer acting on a tensor-encoded configuration. The
executor is not trained; its weights are \emph{constructed} to compute Definition~\ref{def:step}
exactly, in arithmetic disjoint from a conventional ALU. This continues the compiled-transformer line
of work~\cite{weiss2021,lindner2023,giannou2023}, which establishes that transformer weights can be set
by hand to execute symbolic programs; our contribution is to make the executed program an arbitrary
compiled binary and to put the resulting exactness to use as an audit oracle.

\begin{definition}[Bipolar encoding]
Let $\beta : \{0,1\} \to \bp,\ \beta(x) = 2x-1$, extended componentwise to bit-vectors. A word
$w\in\Word$ with binary digits $w_0,\dots,w_{31}$ is encoded as
$\mathbf{b}(w) = \big(\beta(w_0),\dots,\beta(w_{31})\big) \in \bp^{32}$, with inverse
$\beta^{-1}(y) = (y{+}1)/2$.
\end{definition}

\begin{definition}[State tensor]
A configuration is encoded as a matrix $S \in \bp^{d \times K}$ whose columns are the machine's
``cells'' (registers, then a memory window). Rows $0\text{--}31$ hold the cell's 32 value bits in
bipolar form; rows $32\text{--}42$ hold a bipolar encoding of the cell's address/position; the
remaining rows carry type and control fields, including the encoded $pc$.
\end{definition}

The layer computes one step in three constructed stages. Opcode dispatch is \emph{branchless}: it
evaluates all $40$ candidate results and selects with a one-hot mask, so the whole forward pass is a
fixed sequence of tensor contractions (and thus a single fused GPU kernel).

\begin{enumerate}[nosep,leftmargin=1.4em]
\item \textbf{Decode and operand fetch.} An attention head reads $I_{pc}$; further heads select the
      operand columns $b,c$. With keys/queries set to the position rows, the score matrix
      $Q K^{\top}$ is maximised exactly at the addressed column, so a hard (or sharp soft) argmax is a
      \emph{one-hot} selector $\pi \in \{0,1\}^{K}$ and $S\pi$ extracts the operand.
\item \textbf{ALU.} A constructed MLP implements each opcode's word function as a Boolean/arithmetic
      circuit over $\bp$. Bitwise ops are affine maps composed with the sign nonlinearity; addition,
      multiplication, shifts, comparisons and division are realised as fixed circuits on the bit rows
      (carry propagation, partial-product accumulation, restoring division), each producing a
      candidate result $r_{\mathit{op}}$.
\item \textbf{Select and write back.} With the decoded opcode as a one-hot vector $e_{\mathit{op}}$,
      the committed result is $r = \sum_{\mathit{op}} (e_{\mathit{op}})\, r_{\mathit{op}}$; the
      $pc$ update is selected the same way (fall-through, jump target, or branch target). A final
      \emph{error-correction} step snaps every value to $\bp$ by $y \mapsto \operatorname{sign}(y)$,
      restoring an exact bipolar codeword and preventing numerical drift across cycles.
\end{enumerate}

\begin{proposition}[One-cycle faithfulness]\label{prop:faithful}
Let $T$ be the constructed layer and let $\mathrm{enc}$ be the state encoding. For every non-terminal
configuration $\sigma$ with $\sigma \step_P \sigma'$,
\[
   T\big(\mathrm{enc}(\sigma)\big) = \mathrm{enc}(\sigma').
\]
\end{proposition}
\begin{proof}[Proof sketch]
Each stage is exact by construction. The position rows make $QK^\top$ diagonally dominant at the
addressed column, so the selector is exactly one-hot and operand fetch is $S\pi = \mathbf b(R(b))$.
Each candidate circuit computes its opcode's word function on $\bp$ bits: e.g.\ the adder reproduces
$\bmod\,2^{32}$ addition by ripple carry, and comparison circuits reproduce $\textsc{Slt}/\textsc{SltU}$.
The one-hot opcode mask commits exactly the intended $r_{\mathit{op}}$ and $pc'$. The final sign step is
idempotent on $\bp$, so the output equals $\mathrm{enc}(\sigma')$ bit-for-bit. Composing the constructed
maps gives the claim.
\end{proof}

\begin{corollary}[Whole-program faithfulness]\label{cor:whole}
Iterating $T$ from $\mathrm{enc}(\sigma_0)$ for $k$ steps encodes $\sigma_0 \step_P^{k} \sigma_k$;
hence if $P$ halts within budget $B$ the transformer's decoded \texttt{a0} equals $\sem{P}(\sigma_0)$.
\end{corollary}

Corollary~\ref{cor:whole} is the sense in which ``running on the transformer'' equals ``running the
program.'' Independence from a CPU is structural: the value domain is $\bp$, the operations are tensor
contractions and a sign nonlinearity, and (on GPU) the physical fault modes are entirely different from
those of a scalar integer pipeline.

\section{Compilation pipeline and the mini Lean runtime}\label{sec:pipeline}

\subsection{From C to NISA}
The front end compiles C to RV32I with an off-the-shelf cross-compiler
(\texttt{riscv64-linux-gnu-gcc}, \texttt{-march=rv32im}, \texttt{-mabi=ilp32}), then translates RV32I to
NISA instruction-by-instruction; a data-section pass lays out globals and materialises them with
self-initialising stores, and \texttt{\%hi/\%lo} relocations are resolved against that layout. Two
scratch registers are reserved for the translation. \emph{This compiler is itself software and can be
wrong}; two miscompilations were found and fixed during development (a global read-modify-write used as
an array index; a scratch-register collision), and the differential method continues to surface codegen
defects (\S\ref{sec:results}). Crucially, a compiler bug here is a bug in the \emph{auditor}: it can
only cause a spurious disagreement, never a false certification, because the audit never emits a proof
(\S\ref{sec:audit}).

\subsection{Executing Lean's emitted C}
Lean compiles a \texttt{Decidable} instance to C against its runtime ABI: values are
\texttt{lean\_object*}, with arbitrary-precision \texttt{Nat}/\texttt{Int}, boxed constructors, and the
\texttt{lean\_nat\_*}/\texttt{lean\_int\_*}/\texttt{lean\_ctor\_*} primitives. We provide a freestanding
\emph{mini runtime} implementing the fragment of this ABI that decision procedures observe:

\begin{definition}[Object model]
A \texttt{lean\_object*} is either a \emph{tagged scalar} --- low bit $1$, value $\lfloor \cdot/2\rfloor$,
used for $\texttt{Nat} < 2^{31}$ --- or a heap pointer (low bit $0$) into a bump arena. A heap
\texttt{Nat} is a little-endian array of base-$2^{32}$ limbs $[n \mid \ell_0 \mid \dots \mid \ell_{n-1}]$;
an \texttt{Int} is $[\,\mathrm{sign} \mid \mathrm{mag}\,]$ with \texttt{mag} a \texttt{Nat} object; a
constructor is $[\,\mathrm{tag} \mid \#\text{fields} \mid f_0 \mid \dots\,]$, with \texttt{List} given by
$\mathrm{nil}=\beta\text{-box}(0)$ and $\mathrm{cons}\,h\,t = \text{ctor}_1\{h,t\}$.
\end{definition}

Only the runtime's own accessors inspect these private layouts, so the model is a drop-in for the ABI
fragment used by decision procedures. The \texttt{Nat} layer (add, sub, mul, schoolbook long division
for div/mod, square-and-multiply pow, bitwise-and, and the comparisons underlying
\texttt{Nat.decEq/decLt/decLe}) is validated end-to-end on the transformer; the \texttt{Int} and
constructor layers are validated against a host reference build. Because every arithmetic routine has an
independent oracle (host \texttt{int}, \texttt{sympy}), the runtime is itself continuously audited by the
same mechanism it enables.

\begin{example}[A compiled decision, three ways]
The goal \texttt{decide (a * b = c)} over \texttt{Nat} compiles to
\texttt{Nat.decEq (Nat.mul a b) c : Bool}. The audit computes this Boolean (i) as Lean's
\texttt{native\_decide} would, (ii) on the transformer via the mini runtime, and (iii) in a pure oracle
($a\cdot b = c$ over $\Z$). For $a=2^{64}{-}59,\ b=2^{64}{-}83$ (a $\sim$128-bit product) all three agree
on \texttt{true}; for the perturbed $c'=c{+}1$ all three agree on \texttt{false}. The false cases matter:
they witness that the check is not vacuously passing.
\end{example}

\section{The audit method}\label{sec:audit}

Fix a decidable predicate with intended semantics $f : X \to \Bool$. Three evaluators approximate $f$:
\[
  \eval_{\mathrm{lean}} \ (\text{compiled } d,\ \text{on CPU}),\qquad
  \eval_{\mathrm{trans}} \ (\text{same } d,\ \text{via runtime}\circ\text{pipeline}\circ\text{transformer}),\qquad
  \eval_{\mathrm{orac}} \ (\text{independent implementation of } f).
\]
\texttt{native\_decide} accepts $P(x)$ precisely when $\eval_{\mathrm{lean}}(x)=\texttt{true}$.

\begin{definition}[Audit outcome]
For a finite test set $X_0\subseteq X$, the audit reports \emph{disagreement} if there is $x\in X_0$
with $\eval_{\mathrm{trans}}(x)\neq \eval_{\mathrm{orac}}(x)$ or with
$\eval_{\mathrm{lean}}(x)\neq\eval_{\mathrm{orac}}(x)$ (when $\eval_{\mathrm{lean}}$ is available on
$x$), and \emph{agreement} otherwise.
\end{definition}

\begin{theorem}[Soundness of the falsifier]\label{thm:sound}
Suppose that on input $x$ the pipeline and transformer are faithful --- i.e.\
$\eval_{\mathrm{trans}}(x) = g(x)$, where $g$ is the mathematical function the compiled procedure
\emph{denotes} --- and that the oracle is correct, $\eval_{\mathrm{orac}}(x)=f(x)$. If
$\eval_{\mathrm{lean}}(x)=\texttt{true}$ while $\eval_{\mathrm{trans}}(x)=\texttt{false}$, then the
compiled procedure Lean executed does not denote $f$ at $x$; consequently the \texttt{native\_decide}
proof of $P(x)$ rests on an unfaithful computation and $P(x)$ is not thereby established.
\end{theorem}
\begin{proof}
By Corollary~\ref{cor:whole} and the hypothesis, $\eval_{\mathrm{trans}}(x)=g(x)=\texttt{false}$, so
$g(x)\neq\texttt{true}$. Lean's acceptance required the compiled evaluation to yield \texttt{true};
since $g(x)=\texttt{false}$, the value \texttt{native\_decide} trusted differs from the denotation $g$
of the very procedure it ran, or $g$ differs from $f$. In either case the chain
\texttt{ofReduceBool} relied upon --- ``the compiled \texttt{decide }$P$ evaluates to \texttt{true},
therefore $P$'' --- is instantiated with a \texttt{true} that does not reflect $f(x)$; the inference to
$P(x)$ is unjustified.
\end{proof}

\begin{theorem}[Incompleteness]\label{thm:incomplete}
Agreement on $X_0$ does not entail $\eval_{\mathrm{lean}}=f$ on $X_0$, let alone on $X$.
\end{theorem}
\begin{proof}
If all three evaluators share a defect $\delta$ at $x$ (e.g.\ the same off-by-one), then
$\eval_{\mathrm{lean}}(x)=\eval_{\mathrm{trans}}(x)=\eval_{\mathrm{orac}}(x)\neq f(x)$ yet the audit
reports agreement. And for $x\notin X_0$ nothing is tested. Hence agreement is consistent with
unsoundness.
\end{proof}

Theorems~\ref{thm:sound}--\ref{thm:incomplete} together say the audit is a \emph{sound but incomplete}
falsifier: it can certify the presence of a fault but never its absence. This is the formal content of
the Dijkstra principle and it dictates the tactics used to enlarge coverage and independence:

\begin{description}[leftmargin=1.2em,itemsep=1pt]
\item[Differential.] Compare $\eval_{\mathrm{trans}}$ against $\eval_{\mathrm{orac}}$ over many inputs~\cite{mckeeman1998,yang2011},
      weighted toward edges ($0,1$, $2^{31},2^{32},2^{63}$, negatives, empty structures) where fast-path
      discrepancies concentrate.
\item[Metamorphic~\cite{chen1998}.] Test relations that must hold for \emph{all} inputs and require no separate oracle,
      e.g.\ $a\cdot b = b\cdot a$, $\gcd(a,b)\cdot\mathrm{lcm}(a,b)=a\cdot b$, or ``factor then multiply
      back.'' The identity is the oracle, closing the shared-oracle-bug gap of Theorem~\ref{thm:incomplete}.
\item[Bounded-exhaustive.] For finite domains, enumerate $X_0=X$; agreement then \emph{does} entail
      faithfulness on $X$.
\item[Redundant execution.] Run the identical compiled binary on a second architecture (aarch64 vs.\
      amd64 under emulation) and across compiler/toolchain versions; deterministic integer results must
      agree bit-for-bit. This is the only defence against the fault class of \S\ref{sec:threat}.
\end{description}

\section{Threat model}\label{sec:threat}

\paragraph{Auditor faults are safe.} A bug anywhere in Transformer-Oracle --- front end, translator, runtime,
or executor --- can only make $\eval_{\mathrm{trans}}$ wrong. By Theorem~\ref{thm:sound} the audit
emits no proof; a faulty auditor yields at worst a spurious disagreement, which is investigated and
attributed. The system is fail-safe by design: it never certifies.

\paragraph{Shared defects.} The residual blind spot of Theorem~\ref{thm:incomplete} is a defect common
to Lean's path and the auditor's path. The mitigations are (i) implementation diversity --- a
schoolbook bignum in the runtime versus GMP-style code in Lean's fast paths versus \texttt{sympy} in
the oracle --- and (ii) metamorphic checks, which do not depend on any single implementation being
correct.

\paragraph{Silent hardware data corruption.} ``Mercurial cores''~\cite{hochschild2021,dixit2021} produce deterministic wrong results on
particular data without any fault signal. Such an event inside a compiled decision procedure is a direct
attack on \texttt{ofReduceBool}: \texttt{native\_decide} would trust a hardware miscomputation.
Redundant execution is the defence, and it is native to this method --- a corruption manifests as a
disagreement between two independent runs or machines, exactly what the differential and
cross-architecture layers detect. (This class threatens all machine-checked proof, since the kernel runs
on the same silicon; \texttt{native\_decide} is merely more exposed, executing more trusted instructions
without re-checking.)

\paragraph{Floating point.} IEEE-754 $+,-,\times,\div,\surd$ are correctly rounded and hence portable and
reproducible; transcendentals (\texttt{sin},\texttt{exp},\texttt{pow},\dots) are \texttt{libm}-defined
and \emph{not} portable. Audits of real-valued \texttt{native\_decide} goals must treat only the former
as cross-platform ground truth, and use interval/ball enclosure~\cite{johansson2017} for the latter.

\section{Case studies and results}\label{sec:results}

\subsection{A concrete \texttt{native\_decide} unsoundness: \texttt{Vector.scanr}}\label{sec:scanr}

The method's flagship result is a confirmed soundness bug in the Batteries library, filed upstream as
issue~\#1924~\cite{batteries1924}. The reference definition of \texttt{Vector.scanrM} is correct, but
its compiled replacement \texttt{@[implemented\_by] scanrMFast} contains an off-by-one: the inner loop
executes \texttt{acc.uset pred cur}, storing the \emph{old} accumulator, where it must store the freshly
computed value (\texttt{acc.uset pred next}). Every entry is therefore computed but written one slot too
late. Concretely, for $\texttt{scanr}\ (\cdot{+}\cdot)\ 0\ [1,2,3]$ the kernel reduces the reference to
the correct $[6,5,3,0]$, while the compiled fast path---the one \texttt{native\_decide} executes---yields
$[5,3,0,0]$.

Because \texttt{decide} reduces the reference definition while \texttt{native\_decide} runs the fast
path, the two disagree on a \emph{decidable equality}, and one can derive \texttt{False} using no axioms
beyond those \texttt{native\_decide} already assumes (\texttt{ofReduceBool}, \texttt{trustCompiler}). This
is exactly the configuration of Theorem~\ref{thm:sound}: an independent evaluation---here the kernel, and
equally the transformer running the reference \texttt{scanr}---exhibits a value the compiled procedure
does not, proving the compiled artefact unfaithful and any \texttt{native\_decide} proof that relies on it
unsound. The defect lives precisely in the \texttt{@[implemented\_by]} trust surface of
\S\ref{sec:background}, which every downstream project inherits.

A systematic differential/metamorphic sweep of the value-reachable
\texttt{@[implemented\_by]}/\texttt{@[extern]} surface of Lean core and Batteries (several thousand cases
over sorting, \texttt{BitVec} arithmetic, \texttt{Nat}/\texttt{Int} bit and division operations, slicing
and indexing, order-sensitive folds and scans, and hash maps/trees) found \texttt{Vector.scanr} to be the
\emph{sole} divergence---consistent with the principle (\S\ref{sec:background}) that only the fast-path
surface, never a pure Lean definition, can produce a \texttt{native\_decide} soundness bug.

\subsection{Further results}

\begin{itemize}[leftmargin=1.2em,itemsep=2pt]
\item \textbf{Compiler bugs found by self-audit.} Running the mini runtime through the pipeline and
      diffing against a host reference exposed two miscompilations (global read-modify-write indexing;
      scratch-register collision), both fixed, and a third still-open miscompilation of register-heavy
      \emph{composed} \texttt{Int} routines. These are the method applied to its own auditor: the
      transformer-vs-host differential is itself a falsifier (\S\ref{sec:audit}).
\item \textbf{Portability of the method.} The same oracle/metamorphic methodology, pointed at unrelated
      mature C libraries, uncovered concrete defects --- for instance an infinite loop in a
      number-theory routine on a degenerate (zero) modulus, root-caused to a missing precondition guard
      in a residue search. That such a method transfers is evidence that the discipline, not the
      artefact under test, does the work.
\end{itemize}

\section{Related work and discussion}

The approach sits between four lines of work: \emph{translation validation}~\cite{pnueli1998} and
differential testing of compilers~\cite{mckeeman1998,yang2011} (with formally verified compilation,
e.g.\ CompCert~\cite{leroy2009}, as the complementary ``prove it once'' alternative); \emph{compiled
transformers}~\cite{weiss2021,lindner2023,giannou2023}, which construct transformer weights to execute
symbolic programs; \emph{interval/ball arithmetic} for rigorous numerics~\cite{johansson2017} (used here
as an oracle for real-valued goals); and the study of silent hardware errors in large
fleets~\cite{hochschild2021,dixit2021}. Its distinctive elements are (i) a weight-constructed transformer as an
\emph{exact}, independent executor with Proposition~\ref{prop:faithful}, and (ii) a formal account
(Theorems~\ref{thm:sound}--\ref{thm:incomplete}) of what a proof-assistant audit does and does not
establish. The honest scope is a semi-decision procedure for unsoundness: it strengthens confidence in
proportion to coverage and independence, and it can \emph{prove} that a particular \texttt{native\_decide}
is unsound, but it cannot certify soundness. For certainty one still needs kernel-level \texttt{decide} or
a \texttt{Prop}-level proof; the value here is in cheaply and independently hunting for the cases where
the fast, trusted path has gone wrong.

\section*{Reproducibility}
The ISA, pipeline, transformer executor, and the Lean-audit toolkit (mini runtime plus a runnable
three-way example) accompany this paper. The worked audit is invoked with
\texttt{python -m transformer\_oracle.lean\_audit.example\_audit}; the practical guide is
\texttt{docs/auditing-lean-native-decide.md}.

\begin{thebibliography}{99}

\bibitem{demoura2021} L.~de~Moura and S.~Ullrich.
  The Lean~4 theorem prover and programming language.
  In \emph{Automated Deduction (CADE~28)}, LNCS 12699, Springer, 2021.

\bibitem{dijkstra1970} E.~W.~Dijkstra.
  Notes on structured programming. Technical Report EWD249, 1970.
  (``Program testing can be used to show the presence of bugs, but never to show their absence.'')

\bibitem{weiss2021} G.~Weiss, Y.~Goldberg, and E.~Yahav.
  Thinking like transformers.
  In \emph{Proc.\ 38th International Conference on Machine Learning (ICML)}, 2021.

\bibitem{lindner2023} D.~Lindner, J.~Kram\'ar, S.~Farquhar, M.~Rahtz, T.~McGrath, and V.~Mikulik.
  Tracr: Compiled transformers as a laboratory for interpretability.
  In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2023.

\bibitem{giannou2023} A.~Giannou, S.~Rajput, J.~Sohn, K.~Lee, J.~D.~Lee, and D.~Papailiopoulos.
  Looped transformers as programmable computers.
  In \emph{Proc.\ 40th International Conference on Machine Learning (ICML)}, 2023.

\bibitem{vaswani2017} A.~Vaswani, N.~Shazeer, N.~Parmar, J.~Uszkoreit, L.~Jones, A.~N.~Gomez,
  \L.~Kaiser, and I.~Polosukhin.
  Attention is all you need.
  In \emph{Advances in Neural Information Processing Systems (NeurIPS)}, 2017.

\bibitem{mckeeman1998} W.~M.~McKeeman.
  Differential testing for software.
  \emph{Digital Technical Journal}, 10(1):100--107, 1998.

\bibitem{yang2011} X.~Yang, Y.~Chen, E.~Eide, and J.~Regehr.
  Finding and understanding bugs in C compilers.
  In \emph{Proc.\ ACM SIGPLAN Conf.\ on Programming Language Design and Implementation (PLDI)}, 2011.

\bibitem{chen1998} T.~Y.~Chen, S.~C.~Cheung, and S.~M.~Yiu.
  Metamorphic testing: A new approach for generating next test cases.
  Technical Report HKUST-CS98-01, Hong Kong University of Science and Technology, 1998.

\bibitem{pnueli1998} A.~Pnueli, M.~Siegel, and E.~Singerman.
  Translation validation.
  In \emph{Tools and Algorithms for the Construction and Analysis of Systems (TACAS)}, LNCS 1384, 1998.

\bibitem{leroy2009} X.~Leroy.
  Formal verification of a realistic compiler.
  \emph{Communications of the ACM}, 52(7):107--115, 2009.

\bibitem{johansson2017} F.~Johansson.
  Arb: Efficient arbitrary-precision midpoint-radius interval arithmetic.
  \emph{IEEE Transactions on Computers}, 66(8):1281--1292, 2017.

\bibitem{hochschild2021} P.~H.~Hochschild, P.~Turner, J.~C.~Mogul, R.~Govindaraju, P.~Ranganathan,
  D.~E.~Culler, and A.~Vahdat.
  Cores that don't count.
  In \emph{Proc.\ Workshop on Hot Topics in Operating Systems (HotOS)}, 2021.

\bibitem{dixit2021} H.~D.~Dixit, S.~Pendharkar, M.~Beadon, C.~Mason, T.~Chakravarthy, B.~Muthiah,
  and S.~Sankar.
  Silent data corruptions at scale.
  arXiv:2102.11245, 2021.

\bibitem{batteries1924} Batteries (Lean~4 standard library extensions), issue~\#1924:
  \texttt{Vector.scanrM} \texttt{@[implemented\_by] scanrMFast} is incorrect (off-by-one);
  \texttt{native\_decide} disagrees with the kernel.
  \url{https://github.com/leanprover-community/batteries/issues/1924}.

\end{thebibliography}

\end{document}
